Where Do Events Come From?

### 2018-08-07 paper.tex
...

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=h,
  caption={A \CEU program that toggles the state of the LED whenever a radio
           packet is received, terminating on a button press, always with the
           LED off.},
  label={lst.ceu},
]
output high/low LED;
input  high/low BUTTON;
input  Packet   RADIO_RECV;
par/or do
    var high/low v = await BUTTON until (v==low);
with
    finalize with
        emit LED(low);
    end
    loop do
        emit LED(high);
        await RADIO_RECV;
        emit LED(low);
        await RADIO_RECV;
    end
end
\end{lstlisting}

Listing~\ref{lst.ceu} illustrates the main characteristics of \CEU, namely
event-driven I/O, lexically-scoped compositions, and synchronous execution.
%
The program toggles the state of the LED whenever a radio packet is received,
terminating on a button press, always with the LED off.
%
The program first declares the \code{LED} output, and the \code{BUTTON} and
\code{RADIO\_RECV} input events (ln~1--3).
The declarations include a payload type, i.e., each event occurrence carries
a value of that type (\code{high/low} is a boolean type).
%
Then, the program uses a \code{par/or} composition (ln~4--16) to run two lines
of execution, aka \emph{trails}, in parallel:
a single-statement trail that waits for a button press before terminating
(ln~5), and an endless loop that toggles the LED on and off whenever a radio
packet is received (ln~10--15).
The \code{finalize} clause (ln~7--9) ensures that, no matter how its enclosing
trail terminates (e.g., from a button press), the LED will be unconditionally
turned off~(ln~8).

All communication between the application and the environment is done through
the \code{await} and \code{emit} primitives, which awaits an input event and
generates an output event, respectively.

The \code{par/or}, which stands for \emph{parallel-or}, is a lexical
composition that terminates as soon as one of the trails terminates, but which
also automatically aborts and finalizes the other trail(s).
%
In the example, when the button is pressed (ln~5), not only the toggling loop
will be aborted (ln~10--15), but the \code{finalize} clause will turn the LED
off unconditionally (ln~8), since its enclosing block goes out of scope.
%
The \code{par/or} is regarded as an \emph{orthogonal preemption
primitive}~\cite{esterel.preemption} because the trails need not to be tweaked
in order to affect each other.

@@

\subsection{Lexically-Scoped Interrupt Service Routines}
\label{sec.isrs}

Interrupts service routines (ISRs) are software entry points that execute in
response to hardware interrupts from peripherals such as timers and GPIOs.
ISRs are at the lowest interface layer between the hardware and software and
are the absolute source of inputs to programs.
%
An ISR starts to execute as soon as the hardware interrupt occurs, suspending
the ongoing program flow abruptly.
Such asynchronous behavior reflects the inherent concurrent nature of
peripherals interacting with the external world.

@@

We propose to extend \CEU with asynchronous ISRs.
However, asynchronous execution confronts the synchronous mindset of \CEU since
the assumption that reactions are atomic no longer holds.
Not only this may lead to race conditions at a fine grain, but may also affect
the ordering of events at a coarse grain: the effect of an earlier event may be
perceived after the effect of a later event.
%
Still, our goal is to preserve the well-behaved interaction between the \CEU
application and the environment even in the presence of asynchronous ISRs.
%Figure~\ref{fig.async} adapts the event loop of Figure~\ref{fig.env}, with the
%original application represented as the synchronous side and the environment as
%the asynchronous side, all inside the same \CEU application.
%
Our idea is to push all subtleties of asynchronous execution into device
drivers, which emit input events to regular synchronous code in \CEU, exactly
as before.

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={Synchronous application that turns the LED on whenever the button
           is pressed and off whenever it is unpressed.},
  label={lst.gpio.sync},
  escapechar=\%,
]
output high/low PIN_13;  // connected to an LED
input  high/low PIN_02;  // connected to a button
#include "gpio.ceu"      // GPIO driver

@@

Listing~\ref{lst.gpio.async}, the asynchronous side, implements the driver for
the output and input events.
%
An \code{output} implementation (ln~4--6) is similar to a parameterized
subroutine: whenever the application invokes \code{emit}, the output body
executes atomically.
\CEU supports inline C between curly braces (ln~3,5) with interpolation to
evaluate \CEU expressions (e.g., \code{@v}).
This allows drivers to take advantage of existing libraries of embedded
toolchains, such as Arduino.
%
In the example, when the driver is included, it configures pin 13 as output
(ln~3) and sets its new state whenever \code{PIN\_13} is emitted (ln~5).
%
An \code{input} event implementation uses an ISR registered with the
\code{spawn async/isr} primitive (ln~16--18), which is automatically invoked
whenever the associated interrupt occurs (e.g., \code{INT0\_vect}).
%
An ISR in \CEU will typically emit an input event inside its body to awake the
synchronous side (ln~17).
However, although the ISR executes asynchronously when the interrupt occurs,
the \code{emit} goes into a queue (to be discussed) and does not interrupt the
ongoing (suspended) reaction on the synchronous side.
%
In the example, the input driver also configures pin 2 to behave as input and
to generate external interrupts on level transitions (ln~11--15).

@@

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={The A/D driver (ln~1--14) is included in the application inside a
           lexically-scoped block (ln~20--24), which turns off all driver
           functionalities automatically on termination.},
  label={lst.adc},
  escapechar=\%,
]
// A/D DRIVER (adc.ceu)

@@

The example in Listing~\ref{lst.adc} illustrates the use of a lexically-scoped
driver for the A/D converter.
Note that the driver (ln~1--14) and application (ln~16--27) are actually in
separate files.
%
The application is a typical periodic sensor sampling loop, but which
designates an explicit \code{do-end} block (ln~20--24) to include the driver
(ln~21) and use it (ln~22--23).
%
The driver specifies a finalization code (ln~6--12) which executes
unconditionally whenever its enclosing block goes out of scope (ln~24).
The finalize completely disables all A/D functionality to save energy.
%
As discussed in the previous section, the finalizer would also execute if
aborted from a \code{par/or} enclosing the driver.
%
Note that the driver interface (ln~3--4) is visible only inside the block and
thus cannot be inadvertently used outside it.

@@

The addition of asynchronous ISRs poses real threats concerning race conditions
since they interrupt programs at arbitrary points.
%
Listing~\ref{lst.usart} illustrates a minimum USART application (ln~13--22)
that consumes incoming bytes (ln~18--21) as they arrive (ln~16).
%
The USART API (ln~1--4) exposes an input event to signal incoming data (ln~3)
and a circular buffer to prevent data loss (ln~4).
The buffer is indispensable because the microcontroller may cope with the USART 
speed.
%
The driver ISR (ln~6--11) simply appends incoming bytes to the end of the
buffer (ln~9) and signals the application (ln~10).
%
As a possible race condition, note that the ISR may fire and append a new byte
to the buffer (ln~9) just before the application clears it (ln~21), in which
case the new byte will be lost forever.

\CEU treats an \code{async/isr} as a block of code that runs in parallel with
the rest of the program (hence the required prefix \code{spawn}).
%
This way, it is clear for the compiler that the accesses to \code{rx\_buf} in
Listing~\ref{lst.usart} may lead to a race condition (ln~9,21), and \CEU raises
a compile-time error.
%
The programmer is responsible to protect concurrent memory accesses with
\code{atomic} blocks, which disables interrupts (supposedly) for a short period
of time.

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={The USART buffer is shared between the ISR and the application,
           resulting in a pontential race condition.
           \emph{(The driver and application code are in separate files.)}},
  label={lst.usart},
]
// USART INTERFACE

input none USART_RX;
var[32*] byte rx_buf;             // `*' = circular

@@

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={The USART driver now exposes a safer (and more friendly) interface
           to the application.
           \emph{(The driver and application code are in separate files.)}},
  label={lst.usart.code},
]
// USART INTERFACE

@@

input none USART_RX;
var[32*] byte rx_buf;

@@

Our runtime
%implements the event loop proposed in Figure~\ref{fig.async}, which
alternates between synchronous, well-behaved execution, and asynchronous,
unpredictable ISRs.
%
Since the language now has full control over the event loop and only executes
from interrupts, it is possible for the runtime to enter in sleep mode
automatically to save energy.

Listing~\ref{lst.env} is a realization of this architecture:
%
The runtime defines an input queue (ln~1) in which ISRs enqueue new events
(e.g., Listing~\ref{lst.usart.code}, ln~12).
%
In the \code{main} function, we first generate the \emph{boot reaction} (ln~3),
which starts the program, spawns the ISRs, and reaches the first \code{await}
statements in the multiple lines of code.
%
Then, the runtime enters an infinite loop (ln~4--11) that queries the input
queue (ln~6) to awake the program (ln~7).
%
Note that ISRs execute asynchronously with the loop, but they typically only
emit an input event into the queue which will be queried in a subsequent
iteration.
%
Note also that if the queue is empty, the runtime enters in sleep mode to save
energy (ln~9), and will only awake on a new hardware interrupt.

Microcontrollers typically support multiple levels of sleep mode, each
progressively saves more energy at the expense of keeping less functionality
active.
%
As an example, the least efficient mode of the \emph{ATmega328P} allows for
timer interrupts since it keeps its internal clock active, while the most
efficient mode turns off all peripherals and can only awake from external
interrupts.

Our runtime supports three compile-time configurations for \code{ceu\_pm\_sleep}
(ln~9):
%
The first configuration is a \emph{nop} which simply keeps the event loop
running all the time without ever sleeping.
This configuration is useful when introducing new platforms.
%
The second configuration chooses a (inefficient) sleep mode that keeps all
peripherals active.
Although this configuration is not the most energy efficient, at least, it
requires no extra assistance from the drivers.
%
The third configuration keeps a bit vector of the active drivers during runtime
and chooses the most efficient mode, querying this vector every time
\code{ceu\_pm\_sleep} is called.

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={Interaction between the power manager, USART driver, and application
           (each code is actually in a separate file).},
  label={lst.pm},
  escapechar=\%,
]
// POWER MANAGER (in C)

@@

The third configuration achieves optimal efficiency but requires a tight
interaction between the drivers and the power management runtime.
%
Listing~\ref{lst.pm} unveils the power manager (ln~1--18), the modified USART
driver (ln~20--29), and an illustrative application (ln~31--43).
%
The application is a loop that awaits in parallel for a button press (ln~36) or
receiving 10 bytes (ln~39).
Let's call this initial state \code{STATE-A}.
The \code{par/or} terminates when either of them occurs, going to the next line
that awaits another button click (ln~42).
Let's call this other state \code{STATE-B}.
After the button click, the program loops back to \code{STATE-A}.
%
While in \code{STATE-A}, the program can awake from USART and external
interrupts, which means that the power manager should choose a sleep mode in
which the USART remains operational.
%
While in \code{STATE-B}, only external interrupts should awake the program,
which means that the power manager may use the most efficient sleep mode.
%
The power manager enumerates all microcontroller's peripherals (ln~3--8) and,
before sleeping, queries their states (ln~11,13) to choose the most appropriate
sleep mode (ln~12,14,16).
%
The USART driver needs to be extended with calls to enable and disable the
USART state inside the power manager: just before awaiting, the driver enables
the USART (ln~24) and creates a finalization block to unregister it on
termination (ln~25--27).
Termination may occur either directly after receiving the requested number of
bytes, or indirectly if the \code{par/or} in the application terminates on a
button click (ln~36).

Note that applications need not to be explicit about energy management to take
advantage of sleep modes.
All happens automatically because of the synchronous-reactive execution model
and energy-aware runtime of \CEU.

@@

We extend the synchronous language \CEU with asynchronous interrupt service
routines (ISRs) to take control of the whole event loop in reactive systems,
from inputs to outputs.
%
ISRs empower reactive applications to also implement their own device drivers
and self-generate inputs, bypassing the need for a host environment.
%
However, asynchronous execution confronts the well-behaved semantics of
synchronous languages with possible race conditions.
%
To mitigate this threat, we extend the static concurrency checks of \CEU to
detect data races between ISRs and the application.

Our approach for writing drivers relies on the lexical structured mechanisms of
\CEU, such as parallel compositions and abortion of lines of execution, to
offer stronger guarantees.
%
For instance, the visibility of drivers can be delimited to scoped blocks to
avoid resource leaks after usage.
%
The same mechanism allows drivers to signal the power manager that associated
peripherals are no longer required, allowing applications to use optimal
sleep modes to save energy automatically.
%
Our initial tests show optimal energy savings for applications in idle states.

>>>>

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=h,
  caption={A \CEU program that blinks the LED whenever a radio
           packet is received, terminating on a button press, always with the
           LED off.},
  label={lst.ceu},
]
output high/low LED;
input  high/low BUTTON;
input  Packet   RADIO_RECV;
par/or do
    var high/low v = await BUTTON until (v==low);
with
    finalize with
        emit LED(low);
    end
    loop do
        emit LED(high);
        await RADIO_RECV;
        emit LED(low);
        await RADIO_RECV;
    end
end
\end{lstlisting}

Listing~\ref{lst.ceu} illustrates the main characteristics of \CEU, namely
event-driven I/O, lexically-scoped compositions, and synchronous execution.
%
The program toggles the state of the LED whenever a radio packet is received,
terminating on a button press, always with the LED off.
%
The program first declares the \code{LED} output, and the \code{BUTTON} and
\code{RADIO\_RECV} input events (ln~1--3).
The declarations include a payload type, i.e., each event occurrence carries
a value of that type (\code{high/low} is a boolean type).
%
Then, the program uses a \code{par/or} composition (ln~4--16) to run two lines
of execution, aka \emph{trails}, in parallel:
a single-statement trail that waits for a button press before terminating
(ln~5), and an endless loop that toggles the LED on and off whenever a radio
packet is received (ln~10--15).
The \code{finalize} clause (ln~7--9) ensures that, no matter how its enclosing
trail terminates (e.g., from a button press), the LED is unconditionally
turned off~(ln~8).

All communication between the application and the environment is done through
the \code{await} and \code{emit} primitives, which awaits an input event and
generates an output event, respectively (ln~5,8,11--14 in the example).

The \code{par/or}, which stands for \emph{parallel-or}, is a lexical
composition that terminates as soon as one of the trails terminates, but which
also automatically aborts and finalizes the other trail(s).
%
In the example, when the button is pressed (ln~5), not only the toggling loop
will be aborted (ln~10--15), but the \code{finalize} clause will turn the LED
off (ln~8) since its enclosing block goes out of scope.
%
The \code{par/or} is regarded as an \emph{orthogonal preemption
primitive}~\cite{esterel.preemption} because the trails need not to be tweaked
to affect each other.

@@

\subsection{Scoped Interrupt Service Routines}
\label{sec.isrs}

Interrupts service routines (ISRs) are software entry points that execute in
response to hardware interrupts from peripherals such as timers and GPIOs.
ISRs are at the lowest interface layer between the hardware and software and
are the absolute source of inputs to programs.
%
An ISR starts to execute as soon as a hardware interrupt occurs, suspending
the ongoing program flow abruptly.
Such asynchronous behavior reflects the inherent concurrent nature of
peripherals interacting with the real world.

@@

We propose to extend \CEU with asynchronous ISRs.
However, asynchronous execution confronts the synchronous mindset of \CEU since
the assumption that reactions are atomic no longer holds.
Not only this might lead to race conditions at a fine grain, but might also
affect the ordering of events at a coarse grain: the effect of an earlier event
might be perceived after the effect of a later event.
%
Still, our goal is to preserve the well-behaved interaction between the \CEU
application and the environment even in the presence of asynchronous ISRs.
%Figure~\ref{fig.async} adapts the event loop of Figure~\ref{fig.env}, with the
%original application represented as the synchronous side and the environment as
%the asynchronous side, all inside the same \CEU application.
%
Our idea is to push all subtleties of asynchronous execution into device
drivers, which emit input events to regular synchronous code in \CEU, exactly
as before.

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={Synchronous application that turns the LED on whenever the button
           is pressed and off whenever it is unpressed.},
  label={lst.gpio.sync},
  escapechar=\%,
]
output high/low PIN_13;  // connected to LED
input  high/low PIN_02;  // connected to button
#include "gpio.ceu"      // GPIO driver

@@

Listing~\ref{lst.gpio.async}, the asynchronous side, implements the driver for
the output and input events.
%
An \code{output} implementation (ln~4--6) is similar to a parameterized
subroutine: whenever the application invokes \code{emit}, the output body
executes atomically.
\CEU supports inline C between curly braces (ln~3,5) with interpolation to
evaluate \CEU expressions (e.g., \code{@v}).
This allows drivers to take advantage of existing libraries of embedded
toolchains such as Arduino.
%
In the example, when the driver is included, it configures pin 13 as output
(ln~3) and sets its new state whenever \code{PIN\_13} is emitted (ln~5).
%
An \code{input} event implementation uses an ISR registered with the
\code{spawn async/isr} primitive (ln~16--18), which is automatically invoked
whenever the associated interrupt occurs (e.g., \code{INT0\_vect}).
%
An ISR in \CEU will typically perform simple operations and emit an input event
inside its body to awake the synchronous side (ln~17).
However, although the ISR executes asynchronously when the interrupt occurs,
the \code{emit} goes into a queue (to be discussed) and does not interrupt the
ongoing (suspended) reaction on the synchronous side.
%
In the example, the input driver also configures pin 2 to behave as input and
to generate external interrupts on level transitions (ln~11--15).

@@

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={The A/D driver (ln~1--14) is included in the application (ln~21)
           inside a lexically-scoped block (ln~20--24), which turns off all
           driver functionalities automatically on termination.
           \emph{(The driver and application are actually in separate files.)}},
  label={lst.adc},
  escapechar=\%,
]
// A/D DRIVER (adc.ceu)

@@

The next example in Listing~\ref{lst.adc} illustrates the use of a
lexically-scoped driver for the A/D converter.
%
The application is a typical periodic sensor sampling loop, but which
designates an explicit \code{do-end} block (ln~20--24) to include the driver
(ln~21) and use it (ln~22--23).
%
The driver specifies a finalization code (ln~6--12) which executes
unconditionally whenever its enclosing block goes out of scope (ln~24).
The finalize completely disables all A/D functionality to save energy
(ln~8--10).
%
As discussed in the previous section, the finalizer would also execute if
aborted from a hypothetical \code{par/or} enclosing the driver.
%
Note that the driver interface (ln~3--4) is visible only inside the block
(ln~20--24) and thus cannot be inadvertently used outside it.

@@

The addition of asynchronous ISRs now poses real threats concerning race conditions
since they interrupt programs at arbitrary points.
%
Listing~\ref{lst.usart} illustrates a minimum USART application (ln~13--22)
that consumes incoming bytes (ln~18--21) as they arrive (ln~16).
%
The USART API (ln~1--4) exposes an input event to signal incoming data (ln~3)
and a buffer to prevent data loss (ln~4).
The buffer is indispensable because the microcontroller might not cope with the USART 
speed.
%
The driver ISR (ln~6--11) simply appends incoming bytes to the end of the
buffer (ln~9) and signals the application (ln~10).
%
As a possible race condition, note that the ISR may fire and append a new byte
to the buffer (ln~9) just before the application clears it (ln~21), in which
case the new byte will be lost forever.

\CEU treats a \code{spawn~async/isr} as a block of code that runs in parallel with
the rest of the program (hence the prefix \code{spawn}).
%
This way, it is clear for the compiler that the accesses to \code{rx\_buf} in
Listing~\ref{lst.usart} may lead to a race condition (ln~9,21), and \CEU raises
a compile-time error.
%
The programmer is responsible to protect concurrent memory accesses with
\code{atomic} blocks, which disables interrupts for a short period
of time (supposedly).

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={The USART buffer is shared between the ISR and the application,
           resulting in a pontential race condition.
           \emph{(The driver and application are actually in separate files.)}},
  label={lst.usart},
]
// USART INTERFACE

input none USART_RX;
var[32] byte rx_buf;

@@

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={The USART driver now exposes a safer (and more friendly) interface
           to the application.
           \emph{(The driver and application are actually in separate files.)}},
  label={lst.usart.code},
]
// USART INTERFACE

@@

input none USART_RX;
var[32] byte rx_buf;

@@

The language runtime
%implements the event loop proposed in Figure~\ref{fig.async}, which
alternates between synchronous, well-behaved execution, and asynchronous,
unpredictable ISRs.
%
Since the language now has full control over the event loop and only executes
from interrupt requests, it is possible for the runtime to enter in sleep mode
automatically to save energy.
%
Listing~\ref{lst.env} is a realization of this architecture:
%
The runtime defines an input queue (ln~1) in which ISRs enqueue new events
(e.g., Listing~\ref{lst.usart.code}, ln~12).
%
In the \code{main} function, we first generate the \emph{boot reaction} (ln~3),
which starts the program, spawns the ISRs, and reaches the first \code{await}
statements in the multiple lines of code of the program.
%
Then, the runtime enters the infinite event loop (ln~4--11) that queries the input
queue (ln~6) to awake the program (ln~7).
%
Note that ISRs execute asynchronously with the loop and
emit an input event into the queue which will be queried in a subsequent
iteration.
%
Note also that if the queue is empty, the runtime enters in sleep mode to save
energy (ln~9), and will only awake on a new hardware interrupt.

Microcontrollers typically support multiple levels of sleep mode, each
progressively saves more energy at the expense of keeping less functionality
active.
%
As an example, the least efficient mode of the
\emph{ATmega328P}~\cite{arduino.atmega328p} allows for
timer interrupts since it keeps its internal clock active, while the most
efficient mode turns off all peripherals and can only awake from external
interrupts (e.g., GPIO falling edge).

Our runtime supports three compile-time configurations for \code{ceu\_pm\_sleep}
(ln~9):
%
The first configuration is a \emph{nop} which simply keeps the event loop
running all the time without ever sleeping.
This configuration is useful when introducing new platforms.
%
The second configuration chooses a (inefficient) sleep mode that keeps all
peripherals active.
Although this configuration is not the most energy efficient, at least, it
requires no extra assistance from the drivers.
%
The third sleep configuration keeps a bit vector of the active drivers during runtime
and chooses the most efficient mode, querying this vector every time
\code{ceu\_pm\_sleep} is called.

\begin{lstlisting}[
  xleftmargin=1em,
  numbers=left,
  basicstyle=\ttfamily\small,
  float=t,
  caption={Interaction between the power manager, USART driver, and application.
           \emph{(Each code is actually in a separate file.)}},
  label={lst.pm},
  escapechar=\%,
]
// POWER MANAGER (in C)

@@

The third configuration achieves optimal efficiency but requires a tight
interaction between the drivers and the power management runtime.
%
Listing~\ref{lst.pm} unveils the power manager (ln~1--18), the modified USART
driver (ln~20--29), and an illustrative application (ln~31--43).
%
The application is a loop that initially awaits in parallel for a button press (ln~36) or
receiving 10 bytes (ln~39).
Let's call this state \code{STATE-A}.
The \code{par/or} terminates when either of them occurs, going to the next line
that awaits another button click (ln~42).
Let's call this other state \code{STATE-B}.
After another button click, the program loops back to \code{STATE-A}.
%
While in \code{STATE-A}, the program can awake from USART and external
interrupts, which means that the power manager should choose a sleep mode in
which the USART remains operational.
%
While in \code{STATE-B}, only external interrupts should awake the program,
which means that the power manager may use the most efficient sleep mode.
%
The power manager enumerates all microcontroller's peripherals (ln~3--8) and,
before sleeping, queries their states (ln~11,13) to choose the most appropriate
sleep mode (ln~12,14,16).
%
The USART driver needs to be extended with calls to enable and disable the
USART state inside the power manager: just before awaiting (ln~28), the driver enables
the USART (ln~24) and creates a finalization block to unregister it on
termination (ln~25--27).
Termination may occur either directly after receiving the requested number of
bytes, or indirectly if the \code{par/or} in the application terminates on a
button click (ln~36).

Note that applications never need to be explicit about energy management to take
advantage of sleep modes.
All happens automatically because of the synchronous reactive execution model
and the interaction between the drivers and energy-aware runtime of \CEU.

@@

We extend the synchronous language \CEU with asynchronous interrupt service
routines (ISRs) to take control of the whole event loop in reactive systems,
from inputs to outputs.
%
ISRs empower reactive applications to also implement their own device drivers
and self-generate inputs, bypassing the need for a host environment.
%
However, asynchronous execution confronts the well-behaved semantics of
synchronous languages with possible race conditions.
%
To mitigate this threat, we extend the static concurrency checks of \CEU to also
detect data races between ISRs and the application.

Our approach for developing drivers relies on the lexical structured mechanisms of
\CEU, such as parallel compositions and abortion of lines of execution, to
offer stronger guarantees.
%
For instance, the visibility of drivers can be delimited to scoped blocks to
avoid resource leaks.
%
The same mechanism allows drivers to signal the power manager that associated
peripherals are no longer required, allowing applications to use optimal
sleep modes automatically.
%
Our initial tests show optimal energy savings for applications in idle states.
